Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-22 PAPER.tex
changed p_alpha's to capital P's

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $p_\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $p_\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = \sum_\alpha p^\alpha \frac{\partial\textbf{C}}{\partial p^\alpha} \equiv \sum_\alpha p^\alpha \textbf{Q}^\alpha
%\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we define $\frac{\partial\textbf{C}}{\partial p^\alpha} \equiv \textbf{Q}^{\alpha}$.
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{p}^{\alpha}  = \sum_\beta ({\F^{-1}})^{\alpha\beta} (\widehat{q}^{\beta} - \widehat{b}^{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \textbf{Q}^{\alpha} \C^{-1} \textbf{Q}^{\beta} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\widehat{q}^\alpha =  \half \textbf{x}^\dagger \invC \textbf{Q}^{\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}^{\alpha} = \half \Tr\left( \mathbf{U} \invC \textbf{Q}^{\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{p}^{\alpha} \rangle = p^\alpha
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}.

@@

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the advantageous behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a modified version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}^{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial p^\alpha$ but in practice also does other things, including implementing a transform of the frequency domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization matrix $\textbf{M}$, the quantity
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}}
\end{equation}
is a sensible estimate of the true power spectrum $\textbf{P}$. To ensure that $\textbf{M}$ correctly normalizes our power spectrum, one may take the expectation value of Equation \eqref{eq:phat} to obtain
\begin{eqnarray}
\langle \widehat{P}^\alpha \rangle &=& \frac{1}{2} \sum_{\beta \gamma} M^{\alpha \gamma}  \textrm{tr} \left( \textbf{R}\textbf{Q}^{\gamma}\textbf{R} \textbf{Q}^{\beta}\right) p^\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right) \nonumber \\
&\equiv & \sum_{\beta} W^{\alpha \beta} p^\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right), \label{eq:Wpplusbias}
\end{eqnarray}
where $W^{\alpha \beta}$ are elements of a window function matrix. Let's consider the first term of this expression (again, we are assuming that the linear bias term is significantly suppressed; and if this is not the case, we are simply assuming that we are setting a conservative upper limit). If $\mathbf{W}$ ends up being the identity matrix for our choices of $\mathbf{R}$ and $\mathbf{M}$, then we recover Equation \eqref{eq:super_unbiased} for the first term, and we have an estimator that has no multiplicative matrix bias. However, Equation \eqref{eq:super_unbiased} is a rather restrictive condition, and it is possible to violate it and still have a sensible (and correctly normalized) power spectrum estimate. In particular, as long as the rows of $\mathbf{W}$ sum to unity, our power spectrum will be correctly normalized. Beyond this, the data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we discuss other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}. In a preview of what is to come, we also stress that the derivation that leads to Equation \eqref{eq:Wpplusbias} assumes that $\mathbf{R}$ and $\mathbf{x}$ are not correlated. If this assumption is violated, a simple application of the (now incorrect) formulae in this section can result in an improperly normalized power spectrum estimator that does not conserve power, i.e., one that has signal loss.
% JEA - is one really free to choose M and R independently?

@@

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{P}_\alpha^\textrm{loss}$ in terms of $\widehat{P}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{P}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{P}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{P}_\alpha^\textrm{loss} \rangle \approx (1- \eta N_f) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N_f) p_\alpha,
\end{equation}
where recall that $N_f$ is the length of $\x$, or the number of frequency channels. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained.

Recalling that $p_\alpha$ is the \emph{true} power spectrum, one sees that when the covariance in the optimal quadratic estimator is naively replaced by an empirical covariance, the resulting power spectrum estimate is biased low, i.e., there is signal loss. This occurs because of couplings between $\widehat{\C}$ and $\x$, which formally means that what was originally a quadratic estimator is no longer quadratic, but contains higher-order correlations. This violates the assumptions implicit in the derivation of $F_{\alpha \alpha}$ as the normalization factor for converting unnormalized bandpowers $\frac{1}{2} \x^t \C^{-1} \Q^{\alpha} \C^{-1} \x$ into properly normalized power spectrum estimates, where the unnormalized bandpowers are assumed to be two-point (i.e., quadratic) statistics \citep{liu_tegmark2011}. The result is an improperly normalized---and thus lossy---power spectrum estimate.

>>>>

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $P^\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $P^\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = \sum_\alpha P^\alpha \frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \sum_\alpha P^\alpha \textbf{Q}^\alpha
%\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we define $\frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \textbf{Q}^{\alpha}$.
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{P}^{\alpha}  = \sum_\beta ({\F^{-1}})^{\alpha\beta} (\widehat{q}^{\beta} - \widehat{b}^{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \textbf{Q}^{\alpha} \C^{-1} \textbf{Q}^{\beta} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\widehat{q}^\alpha =  \half \textbf{x}^\dagger \invC \textbf{Q}^{\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}^{\alpha} = \half \Tr\left( \mathbf{U} \invC \textbf{Q}^{\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{P}^{\alpha} \rangle = P^\alpha
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}.

@@

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the advantageous behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a modified version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}^{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial P^\alpha$ but in practice also does other things, including implementing a transform of the frequency domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization matrix $\textbf{M}$, the quantity
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}}
\end{equation}
is a sensible estimate of the true power spectrum $\textbf{P}$. To ensure that $\textbf{M}$ correctly normalizes our power spectrum, one may take the expectation value of Equation \eqref{eq:phat} to obtain
\begin{eqnarray}
\langle \widehat{P}^\alpha \rangle &=& \frac{1}{2} \sum_{\beta \gamma} M^{\alpha \gamma}  \textrm{tr} \left( \textbf{R}\textbf{Q}^{\gamma}\textbf{R} \textbf{Q}^{\beta}\right) P^\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right) \nonumber \\
&\equiv & \sum_{\beta} W^{\alpha \beta} P^\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right), \label{eq:Wpplusbias}
\end{eqnarray}
where $W^{\alpha \beta}$ are elements of a window function matrix. Let's consider the first term of this expression (again, we are assuming that the linear bias term is significantly suppressed; and if this is not the case, we are simply assuming that we are setting a conservative upper limit). If $\mathbf{W}$ ends up being the identity matrix for our choices of $\mathbf{R}$ and $\mathbf{M}$, then we recover Equation \eqref{eq:super_unbiased} for the first term, and we have an estimator that has no multiplicative matrix bias. However, Equation \eqref{eq:super_unbiased} is a rather restrictive condition, and it is possible to violate it and still have a sensible (and correctly normalized) power spectrum estimate. In particular, as long as the rows of $\mathbf{W}$ sum to unity, our power spectrum will be correctly normalized. Beyond this, the data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we discuss other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}. In a preview of what is to come, we also stress that the derivation that leads to Equation \eqref{eq:Wpplusbias} assumes that $\mathbf{R}$ and $\mathbf{x}$ are not correlated. If this assumption is violated, a simple application of the (now incorrect) formulae in this section can result in an improperly normalized power spectrum estimator that does not conserve power, i.e., one that has signal loss.
% JEA - is one really free to choose M and R independently?

@@

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{P}_\alpha^\textrm{loss}$ in terms of $\widehat{P}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{P}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{P}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $P_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{P}_\alpha^\textrm{loss} \rangle \approx (1- \eta N_f) P_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N_f) P_\alpha,
\end{equation}
where recall that $N_f$ is the length of $\x$, or the number of frequency channels. In the last step we dropped the final term, since it scales as $\eta P_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained.

Recalling that $P_\alpha$ is the \emph{true} power spectrum, one sees that when the covariance in the optimal quadratic estimator is naively replaced by an empirical covariance, the resulting power spectrum estimate is biased low, i.e., there is signal loss. This occurs because of couplings between $\widehat{\C}$ and $\x$, which formally means that what was originally a quadratic estimator is no longer quadratic, but contains higher-order correlations. This violates the assumptions implicit in the derivation of $F_{\alpha \alpha}$ as the normalization factor for converting unnormalized bandpowers $\frac{1}{2} \x^t \C^{-1} \Q^{\alpha} \C^{-1} \x$ into properly normalized power spectrum estimates, where the unnormalized bandpowers are assumed to be two-point (i.e., quadratic) statistics \citep{liu_tegmark2011}. The result is an improperly normalized---and thus lossy---power spectrum estimate.
